---
title: "How Ecological Conditions Shape Human Detection of Birds"
author: "Anushka K. Gupta and Chandra Earl"
date: "`r Sys.Date()`"
output: html_document
---

This RMarkdown is to document the code used....

The environmental variables we are testing are:

1. 2D wind speed [DP1.00001.001] (Summarized in Summary weather statistics [DP4.00001.001])
2. Triple aspirated air temperature [DP1.00003.001] (Summarized in Summary weather statistics [DP4.00001.001])
3. Barometric pressure [DP1.00004.001] (Summarized in Summary weather statistics [DP4.00001.001])
4. Shortwave radiation (primary pyranometer) (DP1.00022.001) (Summarized in Summary weather statistics [DP4.00001.001])
5. Relative humidity (DP1.00098.001) (Summarized in Summary weather statistics [DP4.00001.001])
*6. Precipitation - weighing gauge [DP1.00044.001]
7. Site management and event reporting [DP1.10111.001] (Planned or unplanned disturbances that have the potential to impact or affect data products)
*8. IR biological temperature [DP1.00005.001]
9. Wet deposition chemical analysis [DP1.00013.001]
10. Stable isotopes in precipitation [DP1.00038.001]
*11. Soil CO2 concentration [DP1.00095.001]
*12. Soil temperature [DP1.00041.001]
*13. Soil water content and water salinity [DP1.00094.001]
*14. Soil heat flux plate [DP1.00040.001]
15. Soil stable isotopes, periodic [DP1.10100.001] (bundled in Soil physical and chemical properties, periodic [DP1.10086.001]) 
16. Soil inorganic nitrogen pools and transformations [DP1.10080.001] (bundled in Soil physical and chemical properties, periodic [DP1.10086.001])
17. Coarse downed wood log survey [DP1.10010.001]
18. Litterfall and fine woody debris production and chemistry [DP1.10033.001]
19. Herbaceous clip harvest [DP1.10023.001]
20. Plant presence and percent cover [DP1.10058.001]
*21. eddy covariance? maybe [DP4.00200.001]


# Setup R Environment

```{r libraries, message=FALSE, warning=FALSE, results="hide"}
library(neonUtilities) 
library(RPresence)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(fuzzyjoin)
library(tibble)
```

# Importing Data

```{r}
# Define NEON sites
large_sites <- c(
  "BARR", "BART", "BONA", "CLBJ", "CPER", "DEJU", "DSNY", "GRSM",
  "GUAN", "HARV", "HEAL", "JERC", "JORN", "KONZ", "MLBS", "MOAB",
  "NIWO", "OAES", "ONAQ", "ORNL", "OSBS", "RMNP", "SJER", "SRER",
  "STEI", "TALL", "TEAK", "UKFS", "UNDE", "WOOD", "WREF", "YELL"
)
small_sites <- c(
  "DELA", "LENO", "TOOL", "SOAP", "STER", "PUUM", "KONA",
  "SERC", "DCFS", "NOGP", "LAJA", "BLAN", "SCBI", "ABBY", "TREE"
)

core_sites <- c(
  "BONA", "CLBJ", "CPER", "GUAN", "HARV",
  "KONZ", "NIWO", "ONAQ", "ORNL", "OSBS",
  "PUUM", "SCBI", "SJER", "SRER", "TALL",
  "TOOL", "UNDE", "WOOD", "WREF", "YELL"
)

core_large_sites <- intersect(core_sites, large_sites)

sites <- core_large_sites

#define years
years = as.character(2015:2023)

# Set up directories
if (!dir.exists("data")) {
  dir.create("data")
}
if (!dir.exists("outputs")) {
  dir.create("outputs")
}
```


## Breeding Landbird Point Counts

https://data.neonscience.org/data-products/DP1.10003.001

This data product contains the quality-controlled, native sampling resolution data from NEON's breeding landbird sampling. Breeding landbirds are defined as “smaller birds (usually exclusive of raptors and upland game birds) not usually associated with aquatic habitats” (Ralph et al. 1993). The breeding landbird point counts product provides records of species identification of all individuals observed during the 6-minute count period, as well as metadata which can be used to model detectability, e.g., weather, distances from observers to birds, and detection methods.

```{r import}
# Path to cached dataset
bird_counts_path <- file.path("data", "bird_counts.RData")

# Load cached data if available; otherwise download and save
if (file.exists(bird_counts_path)) {
  load(bird_counts_path)
} else {
  bird.counts <- loadByProduct(
    dpID    = "DP1.10003.001",
    package = "basic",
    release = "RELEASE-2025",
    token   = Sys.getenv("NEON_TOKEN"),
    site    = sites,
    check.size = FALSE
  )
  
  save(bird.counts, file = bird_counts_path)
}
```

## Summary weather statistics

https://data.neonscience.org/data-products/DP4.00001.001

```{r}
# Path to cached dataset
summary_weather_path <- file.path("data", "summary_weather.RData")

# Load cached data if available; otherwise download and save
if (file.exists(summary_weather_path)) {
  load(summary_weather_path)
} else {
  summary.weather <- loadByProduct(
      dpID    = "DP4.00001.001",
      package = "basic",
      release = "RELEASE-2025",
      token   = Sys.getenv("NEON_TOKEN"),
      site    = sites,
      check.size = FALSE
  )
  
  save(summary.weather, file = summary_weather_path)
}
```

## Primary precipitation
https://data.neonscience.org/data-products/DP1.00006.001

```{r eval=FALSE, include=FALSE}
primary_precipitation_path <- file.path("data", "primary_precipitation.RData")
if (file.exists(primary_precipitation_path)) {
  load(primary_precipitation_path)
} else {
  primary.precipitation <- loadByProduct(
    dpID = "DP1.00006.001",
    package = "basic",
    release = "RELEASE-2025",
    token = Sys.getenv("NEON_TOKEN"),
    site = sites,
    check.size = TRUE
  )
  save(primary.precipitation, file = primary_precipitation_path)
}
```

## Site management and event reporting
https://data.neonscience.org/data-products/DP1.10111.001

```{r}
site_management_event_reporting_path <- file.path("data", "site_management_event_reporting.RData")
if (file.exists(site_management_event_reporting_path)) {
  load(site_management_event_reporting_path)
} else {
  site.management.event.reporting <- loadByProduct(
    dpID = "DP1.10111.001",
    package = "basic",
    release = "RELEASE-2025",
    token = Sys.getenv("NEON_TOKEN"),
    site = sites,
    check.size = FALSE
  )
  save(site.management.event.reporting, file = site_management_event_reporting_path)
}
```

## IR biological temperature
https://data.neonscience.org/data-products/DP1.00005.001

```{r eval=FALSE, include=FALSE}
ir_biological_temperature_path <- file.path("data", "ir_biological_temperature.RData")
if (file.exists(ir_biological_temperature_path)) {
  load(ir_biological_temperature_path)
} else {
  ir.biological.temperature <- loadByProduct(
    dpID = "DP1.00005.001",
    package = "basic",
    release = "RELEASE-2025",
    token = Sys.getenv("NEON_TOKEN"),
    site = sites,
    check.size = FALSE
  )
  save(ir.biological.temperature, file = ir_biological_temperature_path)
}
```


## Wet deposition chemical analysis
https://data.neonscience.org/data-products/DP1.00013.001

```{r}
wet_deposition_chemical_analysis_path <- file.path("data", "wet_deposition_chemical_analysis.RData")
if (file.exists(wet_deposition_chemical_analysis_path)) {
  load(wet_deposition_chemical_analysis_path)
} else {
  wet.deposition.chemical.analysis <- loadByProduct(
    dpID = "DP1.00013.001",
    package = "basic",
    release = "RELEASE-2025",
    token = Sys.getenv("NEON_TOKEN"),
    site = sites,
    check.size = FALSE
  )
  save(wet.deposition.chemical.analysis, file = wet_deposition_chemical_analysis_path)
}
```

## Stable isotopes in precipitation
https://data.neonscience.org/data-products/DP1.00038.001

```{r}
stable_isotopes_precipitation_path <- file.path("data", "stable_isotopes_precipitation.RData")
if (file.exists(stable_isotopes_precipitation_path)) {
  load(stable_isotopes_precipitation_path)
} else {
  stable.isotopes.precipitation <- loadByProduct(
    dpID = "DP1.00038.001",
    package = "basic",
    release = "RELEASE-2025",
    token = Sys.getenv("NEON_TOKEN"),
    site = sites,
    check.size = FALSE
  )
  save(stable.isotopes.precipitation, file = stable_isotopes_precipitation_path)
}
```

## Soil CO2 concentration
https://data.neonscience.org/data-products/DP1.00095.001

```{r eval=FALSE, include=FALSE}
soil_co2_concentration_path <- file.path("data", "soil_co2_concentration.RData")
if (file.exists(soil_co2_concentration_path)) {
  load(soil_co2_concentration_path)
} else {
  soil.co2.concentration <- loadByProduct(
    dpID = "DP1.00095.001",
    package = "basic",
    release = "RELEASE-2025",
    token = Sys.getenv("NEON_TOKEN"),
    site = sites,
    check.size = FALSE
  )
  save(soil.co2.concentration, file = soil_co2_concentration_path)
}
```

## Soil temperature
https://data.neonscience.org/data-products/DP1.00041.001

```{r eval=FALSE, include=FALSE}
soil_temperature_path <- file.path("data", "soil_temperature.RData")
if (file.exists(soil_temperature_path)) {
  load(soil_temperature_path)
} else {
  soil.temperature <- loadByProduct(
    dpID = "DP1.00041.001",
    package = "basic",
    release = "RELEASE-2025",
    token = Sys.getenv("NEON_TOKEN"),
    site = sites,
    check.size = TRUE
  )
  save(soil.temperature, file = soil_temperature_path)
}
```

## Soil water content and water salinity
https://data.neonscience.org/data-products/DP1.00094.001

```{r eval=FALSE, include=FALSE}
soil_water_content_salinity_path <- file.path("data", "soil_water_content_salinity.RData")
if (file.exists(soil_water_content_salinity_path)) {
  load(soil_water_content_salinity_path)
} else {
  soil.water.content.salinity <- loadByProduct(
    dpID = "DP1.00094.001",
    package = "basic",
    release = "RELEASE-2025",
    token = Sys.getenv("NEON_TOKEN"),
    site = sites,
    check.size = TRUE
  )
  save(soil.water.content.salinity, file = soil_water_content_salinity_path)
}
```

## Soil heat flux plate
https://data.neonscience.org/data-products/DP1.00040.001

```{r eval=FALSE, include=FALSE}
soil_heat_flux_plate_path <- file.path("data", "soil_heat_flux_plate.RData")
if (file.exists(soil_heat_flux_plate_path)) {
  load(soil_heat_flux_plate_path)
} else {
  soil.heat.flux.plate <- loadByProduct(
    dpID = "DP1.00040.001",
    package = "basic",
    release = "RELEASE-2025",
    token = Sys.getenv("NEON_TOKEN"),
    site = sites,
    check.size = TRUE
  )
  save(soil.heat.flux.plate, file = soil_heat_flux_plate_path)
}
```

## Soil physical and chemical properties, periodic
https://data.neonscience.org/data-products/DP1.10086.001

```{r}
soil_physical_chemical_properties_periodic_path <- file.path("data", "soil_physical_chemical_properties_periodic.RData")
if (file.exists(soil_physical_chemical_properties_periodic_path)) {
  load(soil_physical_chemical_properties_periodic_path)
} else {
  soil.physical.chemical.properties.periodic <- loadByProduct(
    dpID = "DP1.10086.001",
    package = "basic",
    release = "RELEASE-2025",
    token = Sys.getenv("NEON_TOKEN"),
    site = sites,
    check.size = FALSE
  )
  save(soil.physical.chemical.properties.periodic, file = soil_physical_chemical_properties_periodic_path)
}
```

## Coarse downed wood log survey
https://data.neonscience.org/data-products/DP1.10010.001

```{r}
coarse_downed_wood_log_survey_path <- file.path("data", "coarse_downed_wood_log_survey.RData")
if (file.exists(coarse_downed_wood_log_survey_path)) {
  load(coarse_downed_wood_log_survey_path)
} else {
  coarse.downed.wood.log.survey <- loadByProduct(
    dpID = "DP1.10010.001",
    package = "basic",
    release = "RELEASE-2025",
    token = Sys.getenv("NEON_TOKEN"),
    site = sites,
    check.size = FALSE
  )
  save(coarse.downed.wood.log.survey, file = coarse_downed_wood_log_survey_path)
}
```

## Litterfall and fine woody debris production and chemistry
https://data.neonscience.org/data-products/DP1.10033.001

```{r}
litterfall_fine_woody_debris_path <- file.path("data", "litterfall_fine_woody_debris.RData")
if (file.exists(litterfall_fine_woody_debris_path)) {
  load(litterfall_fine_woody_debris_path)
} else {
  litterfall.fine.woody.debris <- loadByProduct(
    dpID = "DP1.10033.001",
    package = "basic",
    release = "RELEASE-2025",
    token = Sys.getenv("NEON_TOKEN"),
    site = sites,
    check.size = FALSE
  )
  save(litterfall.fine.woody.debris, file = litterfall_fine_woody_debris_path)
}
```

## Herbaceous clip harvest
https://data.neonscience.org/data-products/DP1.10023.001

```{r}
herbaceous_clip_harvest_path <- file.path("data", "herbaceous_clip_harvest.RData")
if (file.exists(herbaceous_clip_harvest_path)) {
  load(herbaceous_clip_harvest_path)
} else {
  herbaceous.clip.harvest <- loadByProduct(
    dpID = "DP1.10023.001",
    package = "basic",
    release = "RELEASE-2025",
    token = Sys.getenv("NEON_TOKEN"),
    site = sites,
    check.size = FALSE
  )
  save(herbaceous.clip.harvest, file = herbaceous_clip_harvest_path)
}
```

## Plant presence and percent cover
https://data.neonscience.org/data-products/DP1.10058.001

```{r}
plant_presence_percent_cover_path <- file.path("data", "plant_presence_percent_cover.RData")
if (file.exists(plant_presence_percent_cover_path)) {
  load(plant_presence_percent_cover_path)
} else {
  plant.presence.percent.cover <- loadByProduct(
    dpID = "DP1.10058.001",
    package = "basic",
    release = "RELEASE-2025",
    token = Sys.getenv("NEON_TOKEN"),
    site = sites,
    check.size = FALSE
  )
  save(plant.presence.percent.cover, file = plant_presence_percent_cover_path)
}
```

# Preparing the Data

## Breeding Landbird Point Counts

NEON provides separate tables for point-count survey metadata (`brd_perpoint`) and the bird counts themselves (`brd_countdata`).

1. Create a unique point-count survey ID

Each point-count survey needs a unique identifier so that we can match the detection data (brd_countdata) with the corresponding per-point metadata (brd_perpoint). We extract the survey year from eventID and then combine the plot, point, year, and bout into a single key called pointSurveyID for each of the tables. Note, because the siteID is already included in the plotID (e.g., BART_012), siteID is not included in the construction of `pointSurveyID`.

```{r uniqueIdentifier}
brd_perpoint_clean <- bird.counts$brd_perpoint %>%
  mutate(year = str_extract(eventID, "\\d{4}"))%>%
  mutate(pointSurveyID = paste(plotID, "point", pointID, year, "bout", boutNumber, sep = "_"))

brd_countdata_clean <- bird.counts$brd_countdata %>%
  mutate(year = str_extract(eventID, "\\d{4}"))%>%
  mutate(pointSurveyID = paste(plotID, "point", pointID, year, "bout", boutNumber, sep = "_"))
```

2. Remove incomplete surveys

The `samplingImpractical` field flags point-count surveys that could not be completed (e.g., weather, access issues). For our purposes, we keep only surveys marked as "OK".

```{r}
brd_perpoint_clean <- brd_perpoint_clean %>%
  filter(samplingImpractical == "OK")
```

3. Keep only species-level records

For this workflow, we only focus on detections identified to the species level. Records that are not identified to species (or have `NA` in `taxonRank`) are dropped.

```{r}
brd_countdata_clean <- brd_countdata_clean %>%
  filter(taxonRank == "species")
```

4. Remove duplicate surveys

Each `pointSurveyID` should correspond to a single point-count survey. Occasionally, duplicate IDs appear, likely due to data entry or processing issues. We identify any duplicated `pointSurveyID` values and remove those surveys from both tables.

```{r}
duplicate_pointSurveyID <- brd_perpoint_clean %>%
  count(pointSurveyID) %>%
  filter(n > 1) %>%
  pull(pointSurveyID)

brd_perpoint_clean <- brd_perpoint_clean %>%
  filter(!pointSurveyID %in% duplicate_pointSurveyID)

brd_countdata_clean <- brd_countdata_clean %>%
  filter(!pointSurveyID %in% duplicate_pointSurveyID)
```

5. Join per-point and count tables

Finally, we join the cleaned detection data (`brd_countdata_clean`) with the cleaned per-point metadata (`brd_perpoint_clean`) using `pointSurveyID` as the primary key. After the join, we drop duplicate columns and keep a single, combined dataset.

```{r}
brd_joineddata_clean <- inner_join(
  brd_countdata_clean,
  brd_perpoint_clean,
  by = "pointSurveyID",
  suffix = c(".count", ".perpoint")
)

#Remove duplicate columns between tables
brd_joineddata_clean <- brd_joineddata_clean %>%
  select(-matches("\\.perpoint$"))
names(brd_joineddata_clean) <- gsub("\\.count$", "", names(brd_joineddata_clean))
```

## Summary Weather Statistics

### 2D wind speed

```{r}
wind_speed <- summary.weather$wss_daily_wind

#Average wind speed across all heights and values
windSpeedMean <- brd_joineddata_clean %>%
  filter(siteID %in% sites,
         year %in% years) %>%
  transmute(siteID, plotID, year, date = as.Date(startDate)) %>%
  left_join(
    wind_speed %>%
      transmute(siteID, date = as.Date(date), wssWindSpeedMean) %>%
      group_by(siteID, date) %>%
      summarise(wssWindSpeedMean = mean(wssWindSpeedMean, na.rm = TRUE), .groups = "drop"),
    by = c("siteID", "date")
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    wssWindSpeedMean = na_if(mean(wssWindSpeedMean, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = wssWindSpeedMean) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

#How gusty or changeable the wind was that day, averaged across the tower
windSpeedHorizontalVariance <- brd_joineddata_clean %>%
  filter(siteID %in% sites,
         year %in% years) %>%
  transmute(siteID, plotID, year, date = as.Date(startDate)) %>%
  left_join(
    wind_speed %>%
      transmute(siteID, date = as.Date(date), wssWindSpeedVariance) %>%
      group_by(siteID, date) %>%
      summarise(wssWindSpeedVariance = mean(wssWindSpeedVariance, na.rm = TRUE), .groups = "drop"),
    by = c("siteID", "date")
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    wssWindSpeedVariance = na_if(mean(wssWindSpeedVariance, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = wssWindSpeedVariance) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

#How different the average wind conditions are from bottom to top of the tower
windSpeedVerticalVariance <- brd_joineddata_clean %>%
  filter(siteID %in% sites,
         year %in% years) %>%
  transmute(siteID, plotID, year, date = as.Date(startDate)) %>%
  left_join(
    wind_speed %>%
      transmute(
        siteID,
        date = as.Date(date),
        verticalPosition,
        wssWindSpeedMean
      ) %>%
      group_by(siteID, date, verticalPosition) %>%
      summarise(mean_vert = mean(wssWindSpeedMean, na.rm = TRUE), .groups = "drop") %>%
      group_by(siteID, date) %>%
      summarise(
        windVar_vert = na_if(var(mean_vert, na.rm = TRUE), NaN),
        .groups = "drop"
      ),
    by = c("siteID", "date")
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    windVar_vert = na_if(mean(windVar_vert, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = windVar_vert) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

```

### Triple aspirated air temperature

```{r}
air_temp <- summary.weather$wss_daily_temp

#Average air temp across all heights and values
airTempMean <- brd_joineddata_clean %>%
  filter(siteID %in% sites,
         year %in% years) %>%
  transmute(siteID, plotID, year, date = as.Date(startDate)) %>%
  left_join(
    air_temp %>%
      transmute(siteID, date = as.Date(date), wssTempTripleMean) %>%
      group_by(siteID, date) %>%
      summarise(wssTempTripleMean = mean(wssTempTripleMean, na.rm = TRUE), .groups = "drop"),
    by = c("siteID", "date")
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    wssTempTripleMean = na_if(mean(wssTempTripleMean, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = wssTempTripleMean) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

#Minimum air temp across all heights and values
airTempMin <- brd_joineddata_clean %>%
  filter(siteID %in% sites,
         year %in% years) %>%
  transmute(siteID, plotID, year, date = as.Date(startDate)) %>%
  left_join(
    air_temp %>%
      transmute(siteID, date = as.Date(date), wssTempTripleMinimum) %>%
      group_by(siteID, date) %>%
      summarise(wssTempTripleMinimum = mean(wssTempTripleMinimum, na.rm = TRUE), .groups = "drop"),
    by = c("siteID", "date")
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    wssTempTripleMinimum = na_if(mean(wssTempTripleMinimum, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = wssTempTripleMinimum) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

#Maximum air temp across all heights and values
airTempMax <- brd_joineddata_clean %>%
  filter(siteID %in% sites,
         year %in% years) %>%
  transmute(siteID, plotID, year, date = as.Date(startDate)) %>%
  left_join(
    air_temp %>%
      transmute(siteID, date = as.Date(date), wssTempTripleMaximum) %>%
      group_by(siteID, date) %>%
      summarise(wssTempTripleMaximum = mean(wssTempTripleMaximum, na.rm = TRUE), .groups = "drop"),
    by = c("siteID", "date")
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    wssTempTripleMaximum = na_if(mean(wssTempTripleMaximum, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = wssTempTripleMaximum) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

#How changeable the air temp was that day, averaged across the tower
airTempHorizontalVariance <- brd_joineddata_clean %>%
  filter(siteID %in% sites,
         year %in% years) %>%
  transmute(siteID, plotID, year, date = as.Date(startDate)) %>%
  left_join(
    air_temp %>%
      transmute(siteID, date = as.Date(date), wssTempTripleVariance) %>%
      group_by(siteID, date) %>%
      summarise(wssTempTripleVariance = mean(wssTempTripleVariance, na.rm = TRUE), .groups = "drop"),
    by = c("siteID", "date")
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    wssTempTripleVariance = na_if(mean(wssTempTripleVariance, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = wssTempTripleVariance) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

```

### Barometric pressure

```{r}
pressure <- summary.weather$wss_daily_pres

#Average pressure across all heights and values
pressureMean <- brd_joineddata_clean %>%
  filter(siteID %in% sites,
         year %in% years) %>%
  transmute(siteID, plotID, year, date = as.Date(startDate)) %>%
  left_join(
    pressure %>%
      transmute(siteID, date = as.Date(date), wssStaPresMean) %>%
      group_by(siteID, date) %>%
      summarise(wssStaPresMean = mean(wssStaPresMean, na.rm = TRUE), .groups = "drop"),
    by = c("siteID", "date")
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    wssStaPresMean = na_if(mean(wssStaPresMean, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = wssStaPresMean) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

```

### Shortwave radiation (primary pyranometer)

```{r}
radiation <- summary.weather$wss_daily_shortRad

#Average radiation across all heights and values
radiationMean <- brd_joineddata_clean %>%
  filter(siteID %in% sites,
         year %in% years) %>%
  transmute(siteID, plotID, year, date = as.Date(startDate)) %>%
  left_join(
    radiation %>%
      transmute(siteID, date = as.Date(date), wssShortRadMean) %>%
      group_by(siteID, date) %>%
      summarise(wssShortRadMean = mean(wssShortRadMean, na.rm = TRUE), .groups = "drop"),
    by = c("siteID", "date")
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    wssShortRadMean = na_if(mean(wssShortRadMean, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = wssShortRadMean) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

#Minimum radiation across all heights and values
radiationMin <- brd_joineddata_clean %>%
  filter(siteID %in% sites,
         year %in% years) %>%
  transmute(siteID, plotID, year, date = as.Date(startDate)) %>%
  left_join(
    radiation %>%
      transmute(siteID, date = as.Date(date), wssShortRadMinimum) %>%
      group_by(siteID, date) %>%
      summarise(wssShortRadMinimum = mean(wssShortRadMinimum, na.rm = TRUE), .groups = "drop"),
    by = c("siteID", "date")
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    wssShortRadMinimum = na_if(mean(wssShortRadMinimum, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = wssShortRadMinimum) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

#Maximum radiation across all heights and values
radiationMax <- brd_joineddata_clean %>%
  filter(siteID %in% sites,
         year %in% years) %>%
  transmute(siteID, plotID, year, date = as.Date(startDate)) %>%
  left_join(
    radiation %>%
      transmute(siteID, date = as.Date(date), wssShortRadMaximum) %>%
      group_by(siteID, date) %>%
      summarise(wssShortRadMaximum = mean(wssShortRadMaximum, na.rm = TRUE), .groups = "drop"),
    by = c("siteID", "date")
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    wssShortRadMaximum = na_if(mean(wssShortRadMaximum, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = wssShortRadMaximum) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

```

### Relative humidity

```{r}
humidity <- summary.weather$wss_daily_humid

#Average humidity across all heights and values
humidityMean <- brd_joineddata_clean %>%
  filter(siteID %in% sites,
         year %in% years) %>%
  transmute(siteID, plotID, year, date = as.Date(startDate)) %>%
  left_join(
    humidity %>%
      transmute(siteID, date = as.Date(date), wssRHMean) %>%
      group_by(siteID, date) %>%
      summarise(wssRHMean = mean(wssRHMean, na.rm = TRUE), .groups = "drop"),
    by = c("siteID", "date")
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    wssRHMean = na_if(mean(wssRHMean, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = wssRHMean) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

#Minimum humidity across all heights and values
humidityMin <- brd_joineddata_clean %>%
  filter(siteID %in% sites,
         year %in% years) %>%
  transmute(siteID, plotID, year, date = as.Date(startDate)) %>%
  left_join(
    humidity %>%
      transmute(siteID, date = as.Date(date), wssRHMinimum) %>%
      group_by(siteID, date) %>%
      summarise(wssRHMinimum = mean(wssRHMinimum, na.rm = TRUE), .groups = "drop"),
    by = c("siteID", "date")
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    wssRHMinimum = na_if(mean(wssRHMinimum, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = wssRHMinimum) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

#Maximum humidity across all heights and values
humidityMax <- brd_joineddata_clean %>%
  filter(siteID %in% sites,
         year %in% years) %>%
  transmute(siteID, plotID, year, date = as.Date(startDate)) %>%
  left_join(
    humidity %>%
      transmute(siteID, date = as.Date(date), wssRHMaximum) %>%
      group_by(siteID, date) %>%
      summarise(wssRHMaximum = mean(wssRHMaximum, na.rm = TRUE), .groups = "drop"),
    by = c("siteID", "date")
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    wssRHMaximum = na_if(mean(wssRHMaximum, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = wssRHMaximum) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

```

### Site management and event reporting

```{r warning=FALSE}
eventData <- site.management.event.reporting$sim_eventData

#eventData <- eventData %>%
#  mutate(eventName = paste(eventType, methodTypeChoice, sep = "_"))

siteManagementEvents <- brd_joineddata_clean %>%
  filter(siteID %in% sites, year %in% years) %>%
  transmute(plotID, year, siteID, t = as.POSIXct(startDate), sid = row_number()) %>%
  left_join(
    eventData %>%
      transmute(siteID, es = as.POSIXct(startDate), ee = as.POSIXct(endDate), eventType),
    by = "siteID"
  ) %>%
  group_by(sid, plotID, year) %>%
  summarise(
    eventType = {
      ev <- unique(eventType[!is.na(eventType) & es <= t & t <= ee])
      if (length(ev) == 0) "NoEvent" else ev[1]
    },
    .groups = "drop"
  ) %>%
  group_by(plotID, year) %>%
  summarise(eventType = first(eventType), .groups = "drop") %>%
  pivot_wider(names_from = year, values_from = eventType) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))
```

### Wet deposition chemical analysis

```{r warning=FALSE}
chem <- wet.deposition.chemical.analysis$wdp_chemLab

precipCalcium <- brd_joineddata_clean %>%
  filter(siteID %in% sites, year %in% years) %>%
  transmute(plotID, year, siteID, t = as.POSIXct(startDate), sid = row_number()) %>%
  left_join(
    chem %>%
      transmute(siteID,
                cs = as.POSIXct(setDate),
                ce = as.POSIXct(collectDate),
                precipCalcium),
    by = "siteID"
  ) %>%
  group_by(sid, plotID, year) %>%
  summarise(
    precipCalcium = {
      x <- precipCalcium[!is.na(precipCalcium) & cs <= t & t <= ce]
      if (length(x) == 0) NA_real_ else x[1]
    },
    .groups = "drop"
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    precipCalcium = na_if(mean(precipCalcium, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = precipCalcium) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

precipMagnesium <- brd_joineddata_clean %>%
  filter(siteID %in% sites, year %in% years) %>%
  transmute(plotID, year, siteID, t = as.POSIXct(startDate), sid = row_number()) %>%
  left_join(
    chem %>%
      transmute(siteID,
                cs = as.POSIXct(setDate),
                ce = as.POSIXct(collectDate),
                precipMagnesium),
    by = "siteID"
  ) %>%
  group_by(sid, plotID, year) %>%
  summarise(
    precipMagnesium = {
      x <- precipMagnesium[!is.na(precipMagnesium) & cs <= t & t <= ce]
      if (length(x) == 0) NA_real_ else x[1]
    },
    .groups = "drop"
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    precipMagnesium = na_if(mean(precipMagnesium, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = precipMagnesium) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

precipPotassium <- brd_joineddata_clean %>%
  filter(siteID %in% sites, year %in% years) %>%
  transmute(plotID, year, siteID, t = as.POSIXct(startDate), sid = row_number()) %>%
  left_join(
    chem %>%
      transmute(siteID,
                cs = as.POSIXct(setDate),
                ce = as.POSIXct(collectDate),
                precipPotassium),
    by = "siteID"
  ) %>%
  group_by(sid, plotID, year) %>%
  summarise(
    precipPotassium = {
      x <- precipPotassium[!is.na(precipPotassium) & cs <= t & t <= ce]
      if (length(x) == 0) NA_real_ else x[1]
    },
    .groups = "drop"
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    precipPotassium = na_if(mean(precipPotassium, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = precipPotassium) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

precipSodium <- brd_joineddata_clean %>%
  filter(siteID %in% sites, year %in% years) %>%
  transmute(plotID, year, siteID, t = as.POSIXct(startDate), sid = row_number()) %>%
  left_join(
    chem %>%
      transmute(siteID,
                cs = as.POSIXct(setDate),
                ce = as.POSIXct(collectDate),
                precipSodium),
    by = "siteID"
  ) %>%
  group_by(sid, plotID, year) %>%
  summarise(
    precipSodium = {
      x <- precipSodium[!is.na(precipSodium) & cs <= t & t <= ce]
      if (length(x) == 0) NA_real_ else x[1]
    },
    .groups = "drop"
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    precipSodium = na_if(mean(precipSodium, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = precipSodium) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

precipAmmonium <- brd_joineddata_clean %>%
  filter(siteID %in% sites, year %in% years) %>%
  transmute(plotID, year, siteID, t = as.POSIXct(startDate), sid = row_number()) %>%
  left_join(
    chem %>%
      transmute(siteID,
                cs = as.POSIXct(setDate),
                ce = as.POSIXct(collectDate),
                precipAmmonium),
    by = "siteID"
  ) %>%
  group_by(sid, plotID, year) %>%
  summarise(
    precipAmmonium = {
      x <- precipAmmonium[!is.na(precipAmmonium) & cs <= t & t <= ce]
      if (length(x) == 0) NA_real_ else x[1]
    },
    .groups = "drop"
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    precipAmmonium = na_if(mean(precipAmmonium, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = precipAmmonium) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

precipNitrate <- brd_joineddata_clean %>%
  filter(siteID %in% sites, year %in% years) %>%
  transmute(plotID, year, siteID, t = as.POSIXct(startDate), sid = row_number()) %>%
  left_join(
    chem %>%
      transmute(siteID,
                cs = as.POSIXct(setDate),
                ce = as.POSIXct(collectDate),
                precipNitrate),
    by = "siteID"
  ) %>%
  group_by(sid, plotID, year) %>%
  summarise(
    precipNitrate = {
      x <- precipNitrate[!is.na(precipNitrate) & cs <= t & t <= ce]
      if (length(x) == 0) NA_real_ else x[1]
    },
    .groups = "drop"
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    precipNitrate = na_if(mean(precipNitrate, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = precipNitrate) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

precipSulfate <- brd_joineddata_clean %>%
  filter(siteID %in% sites, year %in% years) %>%
  transmute(plotID, year, siteID, t = as.POSIXct(startDate), sid = row_number()) %>%
  left_join(
    chem %>%
      transmute(siteID,
                cs = as.POSIXct(setDate),
                ce = as.POSIXct(collectDate),
                precipSulfate),
    by = "siteID"
  ) %>%
  group_by(sid, plotID, year) %>%
  summarise(
    precipSulfate = {
      x <- precipSulfate[!is.na(precipSulfate) & cs <= t & t <= ce]
      if (length(x) == 0) NA_real_ else x[1]
    },
    .groups = "drop"
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    precipSulfate = na_if(mean(precipSulfate, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = precipSulfate) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

precipPhosphate <- brd_joineddata_clean %>%
  filter(siteID %in% sites, year %in% years) %>%
  transmute(plotID, year, siteID, t = as.POSIXct(startDate), sid = row_number()) %>%
  left_join(
    chem %>%
      transmute(siteID,
                cs = as.POSIXct(setDate),
                ce = as.POSIXct(collectDate),
                precipPhosphate),
    by = "siteID"
  ) %>%
  group_by(sid, plotID, year) %>%
  summarise(
    precipPhosphate = {
      x <- precipPhosphate[!is.na(precipPhosphate) & cs <= t & t <= ce]
      if (length(x) == 0) NA_real_ else x[1]
    },
    .groups = "drop"
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    precipPhosphate = na_if(mean(precipPhosphate, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = precipPhosphate) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

precipChloride <- brd_joineddata_clean %>%
  filter(siteID %in% sites, year %in% years) %>%
  transmute(plotID, year, siteID, t = as.POSIXct(startDate), sid = row_number()) %>%
  left_join(
    chem %>%
      transmute(siteID,
                cs = as.POSIXct(setDate),
                ce = as.POSIXct(collectDate),
                precipChloride),
    by = "siteID"
  ) %>%
  group_by(sid, plotID, year) %>%
  summarise(
    precipChloride = {
      x <- precipChloride[!is.na(precipChloride) & cs <= t & t <= ce]
      if (length(x) == 0) NA_real_ else x[1]
    },
    .groups = "drop"
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    precipChloride = na_if(mean(precipChloride, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = precipChloride) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

precipBromide <- brd_joineddata_clean %>%
  filter(siteID %in% sites, year %in% years) %>%
  transmute(plotID, year, siteID, t = as.POSIXct(startDate), sid = row_number()) %>%
  left_join(
    chem %>%
      transmute(siteID,
                cs = as.POSIXct(setDate),
                ce = as.POSIXct(collectDate),
                precipBromide),
    by = "siteID"
  ) %>%
  group_by(sid, plotID, year) %>%
  summarise(
    precipBromide = {
      x <- precipBromide[!is.na(precipBromide) & cs <= t & t <= ce]
      if (length(x) == 0) NA_real_ else x[1]
    },
    .groups = "drop"
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    precipBromide = na_if(mean(precipBromide, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = precipBromide) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

pH <- brd_joineddata_clean %>%
  filter(siteID %in% sites, year %in% years) %>%
  transmute(plotID, year, siteID, t = as.POSIXct(startDate), sid = row_number()) %>%
  left_join(
    chem %>%
      transmute(siteID,
                cs = as.POSIXct(setDate),
                ce = as.POSIXct(collectDate),
                pH),
    by = "siteID"
  ) %>%
  group_by(sid, plotID, year) %>%
  summarise(
    pH = {
      x <- pH[!is.na(pH) & cs <= t & t <= ce]
      if (length(x) == 0) NA_real_ else x[1]
    },
    .groups = "drop"
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    pH = na_if(mean(pH, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = pH) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

precipConductivity <- brd_joineddata_clean %>%
  filter(siteID %in% sites, year %in% years) %>%
  transmute(plotID, year, siteID, t = as.POSIXct(startDate), sid = row_number()) %>%
  left_join(
    chem %>%
      transmute(siteID,
                cs = as.POSIXct(setDate),
                ce = as.POSIXct(collectDate),
                precipConductivity),
    by = "siteID"
  ) %>%
  group_by(sid, plotID, year) %>%
  summarise(
    precipConductivity = {
      x <- precipConductivity[!is.na(precipConductivity) & cs <= t & t <= ce]
      if (length(x) == 0) NA_real_ else x[1]
    },
    .groups = "drop"
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    precipConductivity = na_if(mean(precipConductivity, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = precipConductivity) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

```

### Stable isotopes in precipitation
```{r warning=FALSE}
chem_isotope <- stable.isotopes.precipitation$wdi_isoPerSample

d18OWater <- brd_joineddata_clean %>%
  filter(siteID %in% sites, year %in% years) %>%
  transmute(plotID, year, siteID, t = as.POSIXct(startDate), sid = row_number()) %>%
  left_join(
    chem_isotope %>%
      transmute(siteID,
                cs = as.POSIXct(setDate),
                ce = as.POSIXct(collectDate),
                d18OWater),
    by = "siteID"
  ) %>%
  group_by(sid, plotID, year) %>%
  summarise(
    d18OWater = {
      x <- d18OWater[!is.na(d18OWater) & cs <= t & t <= ce]
      if (length(x) == 0) NA_real_ else x[1]
    },
    .groups = "drop"
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    d18OWater = na_if(mean(d18OWater, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = d18OWater) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

d2HWater <- brd_joineddata_clean %>%
  filter(siteID %in% sites, year %in% years) %>%
  transmute(plotID, year, siteID, t = as.POSIXct(startDate), sid = row_number()) %>%
  left_join(
    chem_isotope %>%
      transmute(siteID,
                cs = as.POSIXct(setDate),
                ce = as.POSIXct(collectDate),
                d2HWater),
    by = "siteID"
  ) %>%
  group_by(sid, plotID, year) %>%
  summarise(
    d2HWater = {
      x <- d2HWater[!is.na(d2HWater) & cs <= t & t <= ce]
      if (length(x) == 0) NA_real_ else x[1]
    },
    .groups = "drop"
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    d2HWater = na_if(mean(d2HWater, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = d2HWater) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))
```

### Soil physical and chemical properties, periodic

```{r warning=FALSE}
soilMoisture <- soil.physical.chemical.properties.periodic$sls_soilMoisture
soilpH <- soil.physical.chemical.properties.periodic$sls_soilpH
soilChemistry <- soil.physical.chemical.properties.periodic$sls_soilChemistry %>%
  mutate(horizon = str_extract(sampleID, "(?<=-)[A-Z](?=-)"))

soil_O <- soilMoisture %>%
  filter(horizon == "O") %>%
  mutate(
    collectDate = as.Date(collectDate),
    sampleYear = as.integer(format(collectDate, "%Y"))
  ) %>%
  group_by(plotID, collectDate, sampleYear) %>%
  summarise(
    soilMoisture_O = mean(soilMoisture, na.rm = TRUE),
    .groups = "drop"
  )

soil_M <- soilMoisture %>%
  filter(horizon == "M") %>%
  mutate(
    collectDate = as.Date(collectDate),
    sampleYear = as.integer(format(collectDate, "%Y"))
  ) %>%
  group_by(plotID, collectDate, sampleYear) %>%
  summarise(
    soilMoisture_M = mean(soilMoisture, na.rm = TRUE),
    .groups = "drop"
  )

soil_pH_O <- soilpH %>%
  filter(horizon == "O") %>%
  mutate(
    collectDate = as.Date(collectDate),
    sampleYear = as.integer(format(collectDate, "%Y"))
  ) %>%
  group_by(plotID, collectDate, sampleYear) %>%
  summarise(
    soilInWaterpH = mean(soilInWaterpH, na.rm = TRUE),
    soilInCaClpH = mean(soilInCaClpH, na.rm = TRUE),
    .groups = "drop"
  )

soil_pH_M <- soilpH %>%
  filter(horizon == "M") %>%
  mutate(
    collectDate = as.Date(collectDate),
    sampleYear = as.integer(format(collectDate, "%Y"))
  ) %>%
  group_by(plotID, collectDate, sampleYear) %>%
  summarise(
    soilInWaterpH = mean(soilInWaterpH, na.rm = TRUE),
    soilInCaClpH = mean(soilInCaClpH, na.rm = TRUE),
    .groups = "drop"
  )

soil_chemistry_O <- soilChemistry %>%
  filter(horizon == "O") %>%
  mutate(
    collectDate = as.Date(collectDate),
    sampleYear = as.integer(format(collectDate, "%Y"))
  ) %>%
  group_by(plotID, collectDate, sampleYear) %>%
  summarise(
    nitrogenPercent = mean(nitrogenPercent, na.rm = TRUE),
    organicCPercent = mean(organicCPercent, na.rm = TRUE),
    CNratio = mean(CNratio, na.rm = TRUE),
    .groups = "drop"
  )

soil_chemistry_M <- soilChemistry %>%
  filter(horizon == "M") %>%
  mutate(
    collectDate = as.Date(collectDate),
    sampleYear = as.integer(format(collectDate, "%Y"))
  ) %>%
  group_by(plotID, collectDate, sampleYear) %>%
  summarise(
    nitrogenPercent = mean(nitrogenPercent, na.rm = TRUE),
    organicCPercent = mean(organicCPercent, na.rm = TRUE),
    CNratio = mean(CNratio, na.rm = TRUE),
    .groups = "drop"
  )

soilMoisture_O <- brd_joineddata_clean %>%
  filter(siteID %in% sites, year %in% years) %>%
  transmute(plotID, year, surveyDate = as.Date(startDate), sid = row_number()) %>%
  left_join(soil_O, by = "plotID") %>%
  mutate(
    d_days = ifelse(sampleYear == year,
                    abs(as.numeric(collectDate - surveyDate)),
                    Inf)
  ) %>%
  group_by(sid, plotID, year) %>%
  slice_min(d_days, with_ties = FALSE) %>%
  summarise(
    soilMoisture_O = if (is.infinite(d_days[1])) NA_real_ else soilMoisture_O[1],
    .groups = "drop"
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    soilMoisture_O = na_if(mean(soilMoisture_O, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = soilMoisture_O) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

soilMoisture_M <- brd_joineddata_clean %>%
  filter(siteID %in% sites, year %in% years) %>%
  transmute(plotID, year, surveyDate = as.Date(startDate), sid = row_number()) %>%
  left_join(soil_M, by = "plotID") %>%
  mutate(
    d_days = ifelse(sampleYear == year,
                    abs(as.numeric(collectDate - surveyDate)),
                    Inf)
  ) %>%
  group_by(sid, plotID, year) %>%
  slice_min(d_days, with_ties = FALSE) %>%
  summarise(
    soilMoisture_M = if (is.infinite(d_days[1])) NA_real_ else soilMoisture_M[1],
    .groups = "drop"
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    soilMoisture_M = na_if(mean(soilMoisture_M, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = soilMoisture_M) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

soilpH_O_Water <- brd_joineddata_clean %>%
  filter(siteID %in% sites, year %in% years) %>%
  transmute(plotID, year, surveyDate = as.Date(startDate), sid = row_number()) %>%
  left_join(soil_pH_O, by = "plotID") %>%
  mutate(
    d_days = ifelse(sampleYear == year,
                    abs(as.numeric(collectDate - surveyDate)),
                    Inf)
  ) %>%
  group_by(sid, plotID, year) %>%
  slice_min(d_days, with_ties = FALSE) %>%
  summarise(
    soilInWaterpH = if (is.infinite(d_days[1])) NA_real_ else soilInWaterpH[1],
    .groups = "drop"
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    soilInWaterpH = na_if(mean(soilInWaterpH, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = soilInWaterpH) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

soilpH_O_CaCl <- brd_joineddata_clean %>%
  filter(siteID %in% sites, year %in% years) %>%
  transmute(plotID, year, surveyDate = as.Date(startDate), sid = row_number()) %>%
  left_join(soil_pH_O, by = "plotID") %>%
  mutate(
    d_days = ifelse(sampleYear == year,
                    abs(as.numeric(collectDate - surveyDate)),
                    Inf)
  ) %>%
  group_by(sid, plotID, year) %>%
  slice_min(d_days, with_ties = FALSE) %>%
  summarise(
    soilInCaClpH = if (is.infinite(d_days[1])) NA_real_ else soilInCaClpH[1],
    .groups = "drop"
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    soilInCaClpH = na_if(mean(soilInCaClpH, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = soilInCaClpH) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

soilpH_M_Water <- brd_joineddata_clean %>%
  filter(siteID %in% sites, year %in% years) %>%
  transmute(plotID, year, surveyDate = as.Date(startDate), sid = row_number()) %>%
  left_join(soil_pH_M, by = "plotID") %>%
  mutate(
    d_days = ifelse(sampleYear == year,
                    abs(as.numeric(collectDate - surveyDate)),
                    Inf)
  ) %>%
  group_by(sid, plotID, year) %>%
  slice_min(d_days, with_ties = FALSE) %>%
  summarise(
    soilInWaterpH = if (is.infinite(d_days[1])) NA_real_ else soilInWaterpH[1],
    .groups = "drop"
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    soilInWaterpH = na_if(mean(soilInWaterpH, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = soilInWaterpH) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

soilpH_M_CaCl <- brd_joineddata_clean %>%
  filter(siteID %in% sites, year %in% years) %>%
  transmute(plotID, year, surveyDate = as.Date(startDate), sid = row_number()) %>%
  left_join(soil_pH_M, by = "plotID") %>%
  mutate(
    d_days = ifelse(sampleYear == year,
                    abs(as.numeric(collectDate - surveyDate)),
                    Inf)
  ) %>%
  group_by(sid, plotID, year) %>%
  slice_min(d_days, with_ties = FALSE) %>%
  summarise(
    soilInCaClpH = if (is.infinite(d_days[1])) NA_real_ else soilInCaClpH[1],
    .groups = "drop"
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    soilInCaClpH = na_if(mean(soilInCaClpH, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = soilInCaClpH) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

soilChemistry_O_percentN <- brd_joineddata_clean %>%
  filter(siteID %in% sites, year %in% years) %>%
  transmute(plotID, year, surveyDate = as.Date(startDate), sid = row_number()) %>%
  left_join(soil_chemistry_O, by = "plotID") %>%
  mutate(
    d_days = ifelse(sampleYear == year,
                    abs(as.numeric(collectDate - surveyDate)),
                    Inf)
  ) %>%
  group_by(sid, plotID, year) %>%
  slice_min(d_days, with_ties = FALSE) %>%
  summarise(
    nitrogenPercent = if (is.infinite(d_days[1])) NA_real_ else nitrogenPercent[1],
    .groups = "drop"
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    nitrogenPercent = na_if(mean(nitrogenPercent, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = nitrogenPercent) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

soilChemistry_M_percentN <- brd_joineddata_clean %>%
  filter(siteID %in% sites, year %in% years) %>%
  transmute(plotID, year, surveyDate = as.Date(startDate), sid = row_number()) %>%
  left_join(soil_chemistry_M, by = "plotID") %>%
  mutate(
    d_days = ifelse(sampleYear == year,
                    abs(as.numeric(collectDate - surveyDate)),
                    Inf)
  ) %>%
  group_by(sid, plotID, year) %>%
  slice_min(d_days, with_ties = FALSE) %>%
  summarise(
    nitrogenPercent = if (is.infinite(d_days[1])) NA_real_ else nitrogenPercent[1],
    .groups = "drop"
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    nitrogenPercent = na_if(mean(nitrogenPercent, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = nitrogenPercent) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

soilChemistry_O_percentC <- brd_joineddata_clean %>%
  filter(siteID %in% sites, year %in% years) %>%
  transmute(plotID, year, surveyDate = as.Date(startDate), sid = row_number()) %>%
  left_join(soil_chemistry_O, by = "plotID") %>%
  mutate(
    d_days = ifelse(sampleYear == year,
                    abs(as.numeric(collectDate - surveyDate)),
                    Inf)
  ) %>%
  group_by(sid, plotID, year) %>%
  slice_min(d_days, with_ties = FALSE) %>%
  summarise(
    organicCPercent = if (is.infinite(d_days[1])) NA_real_ else organicCPercent[1],
    .groups = "drop"
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    organicCPercent = na_if(mean(organicCPercent, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = organicCPercent) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

soilChemistry_M_percentC <- brd_joineddata_clean %>%
  filter(siteID %in% sites, year %in% years) %>%
  transmute(plotID, year, surveyDate = as.Date(startDate), sid = row_number()) %>%
  left_join(soil_chemistry_M, by = "plotID") %>%
  mutate(
    d_days = ifelse(sampleYear == year,
                    abs(as.numeric(collectDate - surveyDate)),
                    Inf)
  ) %>%
  group_by(sid, plotID, year) %>%
  slice_min(d_days, with_ties = FALSE) %>%
  summarise(
    organicCPercent = if (is.infinite(d_days[1])) NA_real_ else organicCPercent[1],
    .groups = "drop"
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    organicCPercent = na_if(mean(organicCPercent, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = organicCPercent) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

soilChemistry_O_CNratio <- brd_joineddata_clean %>%
  filter(siteID %in% sites, year %in% years) %>%
  transmute(plotID, year, surveyDate = as.Date(startDate), sid = row_number()) %>%
  left_join(soil_chemistry_O, by = "plotID") %>%
  mutate(
    d_days = ifelse(sampleYear == year,
                    abs(as.numeric(collectDate - surveyDate)),
                    Inf)
  ) %>%
  group_by(sid, plotID, year) %>%
  slice_min(d_days, with_ties = FALSE) %>%
  summarise(
    CNratio = if (is.infinite(d_days[1])) NA_real_ else CNratio[1],
    .groups = "drop"
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    CNratio = na_if(mean(CNratio, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = CNratio) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

soilChemistry_M_CNratio <- brd_joineddata_clean %>%
  filter(siteID %in% sites, year %in% years) %>%
  transmute(plotID, year, surveyDate = as.Date(startDate), sid = row_number()) %>%
  left_join(soil_chemistry_M, by = "plotID") %>%
  mutate(
    d_days = ifelse(sampleYear == year,
                    abs(as.numeric(collectDate - surveyDate)),
                    Inf)
  ) %>%
  group_by(sid, plotID, year) %>%
  slice_min(d_days, with_ties = FALSE) %>%
  summarise(
    CNratio = if (is.infinite(d_days[1])) NA_real_ else CNratio[1],
    .groups = "drop"
  ) %>%
  group_by(plotID, year) %>%
  summarise(
    CNratio = na_if(mean(CNratio, na.rm = TRUE), NaN),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = year, values_from = CNratio) %>%
  tibble::column_to_rownames("plotID") %>%
  select(all_of(sort(names(.))))

```

## Covariate Data Frame

```{r}
# create the surveycov dataframe
surveycov <-  data.frame(
  windSpeedMean = unlist(windSpeedMean),
  windSpeedHorizontalVariance = unlist(windSpeedHorizontalVariance),
  windSpeedVerticalVariance = unlist(windSpeedVerticalVariance),
  airTempMean = unlist(airTempMean),
  airTempMin = unlist(airTempMin),
  airTempMax = unlist(airTempMax),
  airTempHorizontalVariance = unlist(airTempHorizontalVariance),
  pressureMean = unlist(pressureMean),
  radiationMean = unlist(radiationMean),
  radiationMin = unlist(radiationMin),
  radiationMax = unlist(radiationMax),
  humidityMean = unlist(humidityMean),
  humidityMin = unlist(humidityMin),
  humidityMax = unlist(humidityMax),
  siteManagementEvents = as.factor(unlist(siteManagementEvents)),
  precipCalcium = unlist(precipCalcium),
  precipMagnesium = unlist(precipMagnesium),
  precipPotassium = unlist(precipPotassium),
  precipSodium = unlist(precipSodium),
  precipAmmonium = unlist(precipAmmonium),
  precipNitrate = unlist(precipNitrate),
  precipSulfate = unlist(precipSulfate),
  precipPhosphate = unlist(precipPhosphate),
  precipChloride = unlist(precipChloride),
  precipBromide = unlist(precipBromide),
  pH = unlist(pH),
  precipConductivity = unlist(precipConductivity),
  d18OWater = unlist(d18OWater),
  d2HWater = unlist(d2HWater),
  soilMoisture_O = unlist(soilMoisture_O),
  soilMoisture_M = unlist(soilMoisture_M),
  soilpH_O_Water = unlist(soilpH_O_Water),
  soilpH_M_Water = unlist(soilpH_M_Water),
  soilpH_O_CaCl = unlist(soilpH_O_CaCl),
  soilpH_M_CaCl = unlist(soilpH_M_CaCl),
  soilChemistry_O_percentN = unlist(soilChemistry_O_percentN),
  soilChemistry_M_percentN = unlist(soilChemistry_M_percentN),
  soilChemistry_O_percentC = unlist(soilChemistry_O_percentC),
  soilChemistry_M_percentC = unlist(soilChemistry_M_percentC),
  soilChemistry_O_CNratio = unlist(soilChemistry_O_CNratio),
  soilChemistry_M_CNratio = unlist(soilChemistry_M_CNratio)
  )

# remove the rownames
row.names(surveycov) <- NULL

head(surveycov)

```

# Analyzing the Data

## Baseline Modeling

We define a function to run a baseline model and pull the results. This allows us to run the same function across all available species. 

```{r include=FALSE}
baseline_model <- function(species_df, brd_joineddata_clean, brd_perpoint_clean, sites, years) {
  
  results <- list()
  detection_histories <- list()
  models <- list()
  
  for (i in seq_len(nrow(species_df))) {
    
    sci_name <- species_df$scientificName[i]
    common_name <- species_df$vernacularName[i]
    
    message(
      "Running species: ",
      sci_name,
      " (",
      common_name,
      ")"
    )
    
    # Filter detection data to this species and the sites
    brd_species <- brd_joineddata_clean %>%
      filter(siteID %in% sites, scientificName == sci_name)
    
    # Get valid surveys (plot-year combos with effort)
    valid_plot_years <- brd_perpoint_clean %>%
      filter(
        siteID %in% sites,
        year %in% years
      ) %>%
      distinct(plotID, year)
    
    # Get detections for this species
    detections <- brd_species %>%
      filter(siteID %in% sites) %>%
      group_by(plotID, year) %>%
      summarize(present = 1, .groups = "drop")
    
    # Build plot-year detection matrix
    plot_by_year <- valid_plot_years %>%
      left_join(detections, by = c("plotID","year")) %>%
      mutate(present = replace_na(present, 0)) %>%
      pivot_wider(
        names_from = year,
        values_from = present
      ) %>%
      select(1, sort(names(.)[-1])) %>%
      arrange(plotID)
    
    # Create Pao object
    birds_pao <- createPao(
      data = plot_by_year[, years],
      unitnames = plot_by_year$plotID,
      title = common_name
    )
    
    # Run occupancy model
    model <- occMod(
      data = birds_pao,
      model = list(psi ~ 1, p ~ 1),
      type = "so"
    )
    
    # Extract estimates
    ests <- print_one_site_estimates(mod = model) %>%
      as.data.frame() %>%
      slice(1:2)
    
    rownames(ests) <- NULL
    
    named_est <- list(
      vernacularName = common_name,
      psi_Estimate = ests$est[1],
      psi_SE       = ests$se[1],
      psi_L95      = ests$lower[1],
      psi_U95      = ests$upper[1],
      p_Estimate   = ests$est[2],
      p_SE         = ests$se[2],
      p_L95        = ests$lower[2],
      p_U95        = ests$upper[2],
      neg2loglike  = model$neg2loglike
    )

    detection_histories[[sci_name]] <- plot_by_year
    results[[sci_name]] <- named_est
    models[[sci_name]] <- model
  }
  
  return(list(
    estimates = results,
    detection_histories = detection_histories,
    models = models
  ))
  
}
```

Run function and save the results.

```{r warning=FALSE}
species_input <- brd_joineddata_clean %>%
  filter(
    siteID %in% sites,
    year %in% years
  ) %>%
  distinct(scientificName, vernacularName) %>%
  arrange(scientificName)

baseline_results_df <- baseline_model(
  species_df = species_input,
  brd_joineddata_clean = brd_joineddata_clean,
  brd_perpoint_clean = brd_perpoint_clean,
  sites = sites,
  years = years
)
```

## Covariate Modeling

Define a function to run survey covariates

```{r include=FALSE}
covariate_model <- function(species_df, brd_joineddata_clean, brd_perpoint_clean, sites, years, surveycov, covtouse) {
  
  results <- list()
  detection_histories <- list()
  models <- list()
  
  for (i in seq_len(nrow(species_df))) {
    
    sci_name <- species_df$scientificName[i]
    common_name <- species_df$vernacularName[i]
    
    message(
      "Running species: ",
      sci_name,
      " (",
      common_name,
      "): ",
      covtouse
    )
    
    # Filter detection data to this species and the sites
    brd_species <- brd_joineddata_clean %>%
      filter(siteID %in% sites, scientificName == sci_name)
    
    # Get valid surveys (plot-year combos with effort)
    valid_plot_years <- brd_perpoint_clean %>%
      filter(
        siteID %in% sites,
        year %in% years
      ) %>%
      distinct(plotID, year)
    
    # Get detections for this species
    detections <- brd_species %>%
      filter(siteID %in% sites) %>%
      group_by(plotID, year) %>%
      summarize(present = 1, .groups = "drop")
    
    # Build plot-year detection matrix
    plot_by_year <- valid_plot_years %>%
      left_join(detections, by = c("plotID","year")) %>%
      mutate(present = replace_na(present, 0)) %>%
      pivot_wider(
        names_from = year,
        values_from = present
      ) %>%
      select(1, sort(names(.)[-1])) %>%
      arrange(plotID)
    
    # Create Pao object
    birds_pao <- createPao(
      data = plot_by_year[, years],
      unitnames = plot_by_year$plotID,
      survcov = surveycov,
      title = common_name
    )
    
    # Run occupancy model
    model <- occMod(
      data = birds_pao,
      model = list(
        psi = ~ 1,
        p   = as.formula(paste("~", covtouse))
      ),
      type = "so"
    )
    
    # Extract estimates
    ests <- print_one_site_estimates(mod = model) %>%
      as.data.frame() %>%
      slice(1:2)
    
    rownames(ests) <- NULL
    
    named_est <- list(
      vernacularName = common_name,
      psi_Estimate = ests$est[1],
      psi_SE       = ests$se[1],
      psi_L95      = ests$lower[1],
      psi_U95      = ests$upper[1],
      neg2loglike  = model$neg2loglike,
      slope        = coef(model, 'p', prob = 0.95)[2, 1, drop = TRUE],
      se           = coef(model, 'p', prob = 0.95)[2, 2, drop = TRUE],
      slopeLowerCI = coef(model, 'p', prob = 0.95)[2, 3, drop = TRUE],
      slopeUpperCI = coef(model, 'p', prob = 0.95)[2, 4, drop = TRUE],
      ci_has_zero  = as.integer(
                        coef(model, "p", prob = 0.95)[2, 3] <= 0 &
                        coef(model, "p", prob = 0.95)[2, 4] >= 0
                      )
    )

    detection_histories[[sci_name]] <- plot_by_year
    results[[sci_name]] <- named_est
    models[[sci_name]] <- model
  }
  
  return(list(
    estimates = results,
    detection_histories = detection_histories,
    models = models
  ))
  
}
```

Run across all covariates

```{r warning=FALSE, include=FALSE}
species_input <- brd_joineddata_clean %>%
  filter(
    siteID %in% sites,
    year %in% years
  ) %>%
  distinct(scientificName, vernacularName) %>%
  arrange(scientificName)

listofcovs <- colnames(surveycov)

covariate_results_df <- list()
for (k in seq_len(length(listofcovs))) {
  cov_name <- listofcovs[k]

  covariate_results_df[[cov_name]] <- covariate_model(
    species_df = species_input,
    brd_joineddata_clean = brd_joineddata_clean,
    brd_perpoint_clean = brd_perpoint_clean,
    sites = sites,
    years = years,
    surveycov = surveycov,
    covtouse = cov_name
  )
}

```

# Analyzing Results

Combine the two results into one object and compare covariate models to baseline for each bird

```{r}
species_names <- names(baseline_results_df$estimates)
cov_names     <- names(covariate_results_df)

combined_results <- setNames(lapply(species_names, function(sp) {

  rows <- list(baseline = baseline_results_df$estimates[[sp]])

  for (cv in cov_names) {
    rows[[cv]] <- covariate_results_df[[cv]]$estimates[[sp]]
  }

  bind_rows(lapply(rows, \(x) as.data.frame(as.list(x))), .id = "model") %>%
    tibble::column_to_rownames("model")

}), species_names)

combined_results <- lapply(combined_results, function(df) {
  baseline_val <- df["baseline", "neg2loglike"]

  df$delta_neg2loglike <- df$neg2loglike - baseline_val
  df
})

```

Clean unusable models

```{r}
#if baseline has a p or psi 95% CI of 0-1, remove it
combined_results <- combined_results[
  !vapply(
    combined_results,
    function(df) {
      (df["baseline", "p_L95"] == 0 & df["baseline", "p_U95"] == 1) |
      (df["baseline", "psi_L95"] == 0 & df["baseline", "psi_U95"] == 1)
    },
    logical(1)
  )
]

#if the baseline psi values are all below 0, there likely was not enough data to make a model
min_psi <- 0.01

combined_results <- combined_results[
  !vapply(
    combined_results,
    function(df) {
      with(df["baseline", ],
           psi_Estimate < min_psi &
           psi_L95      < min_psi &
           psi_U95      < min_psi)
    },
    logical(1)
  )
]

#remove covariate models from birds that have NA for se, slopeLowerCI and slopeUpperCI
combined_results <- lapply(combined_results, function(df) {
  keep <- !(is.na(df$se) & is.na(df$slopeLowerCI) & is.na(df$slopeUpperCI))
  keep[rownames(df) == "baseline"] <- TRUE
  df[keep, , drop = FALSE]
})

#Remove species for which there is very little variation in detections at plots. These are valid, though aren't useful for detection since it will mostly always be 1 or close to it

mostly_thresh <- 0.85   # "mostly 1s" within plot
plot_frac_thresh <- 0.9  # >50% of eligible plots

problem_species <- names(covariate_results_df$humidityMax$detection_histories)[
  vapply(covariate_results_df$humidityMax$detection_histories, function(dh) {

    y <- dh[, setdiff(names(dh), "plotID"), drop = FALSE]
    y <- as.matrix(sapply(y, function(x) as.numeric(as.character(x))))

    has_det <- rowSums(y == 1, na.rm = TRUE) >= 1
    if (!any(has_det)) return(FALSE)

    prop1 <- rowMeans(y == 1, na.rm = TRUE)

    mean(prop1[has_det] >= mostly_thresh, na.rm = TRUE) > plot_frac_thresh
  }, logical(1))
]

combined_results <- combined_results[
  !names(combined_results) %in% problem_species
]

#Manually remove
combined_results <- combined_results[
  !names(combined_results) %in% c("Chamaea fasciata",  #this has only one detection ever
                                  "Deltarhynchus flammulatus") #this has only one detection ever
]


```


Summarize across all birds for each covariate

```{r}
#tolerance threshold for zero
eps <- 1

long <- bind_rows(lapply(names(combined_results), function(sp) {
  combined_results[[sp]] %>%
    rownames_to_column("covariate") %>%
    mutate(species = sp)
}))

cov_summary <- long %>%
  filter(covariate != "baseline") %>%
  group_by(covariate) %>%
  summarise(
    n_species = n_distinct(species),

    n_delta_lt0 = sum(delta_neg2loglike < -eps, na.rm = TRUE),
    n_delta_gt0 = sum(delta_neg2loglike >  eps, na.rm = TRUE),
    n_delta_eq0 = sum(abs(delta_neg2loglike) <= eps, na.rm = TRUE),
    avg_delta   = mean(delta_neg2loglike, na.rm = TRUE),

    n_slope_pos = sum(slope > 0, na.rm = TRUE),
    n_slope_neg = sum(slope < 0, na.rm = TRUE),
    avg_slope   = mean(slope, na.rm = TRUE),

    n_CI_incl0  = sum(ci_has_zero, na.rm = TRUE),
    avg_CI      = mean(slopeUpperCI - slopeLowerCI, na.rm = TRUE),

    .groups = "drop"
  )
```

Which species has a slope CI that does not include 0 (and therefore may be significant) for a covariate?
```{r}
covtoplot <- "windSpeedHorizontalVariance"

species_to_plot <- names(combined_results)[
  vapply(combined_results, function(df) {
    covtoplot %in% rownames(df) &&
      df[covtoplot, "ci_has_zero"] == 0
  }, logical(1))
]

species_to_plot

```

Which species have the greatest slopes?
```{r}
covtoplot <- "windSpeedMean"
paramplot <- "slope"

vals <- sapply(names(combined_results), function(sp) combined_results[[sp]][covtoplot, paramplot])
names(vals) <- names(combined_results)

sort(vals, decreasing = FALSE)[1:9]
species_to_plot <- names(sort(vals, decreasing = FALSE)[1:9])

as.character(species_to_plot)
```


# Plotting Results

Let's plot something - plot predicted detection at different covariate values for a species

```{r}
covtoplot <- "windSpeedMean"

# choose which 9 to view (change start index: 1, 10, 19, 28, ...)
start <- 1
sp_batch <- species_to_plot[start:min(start + 8, length(species_to_plot))]
#sp_batch <- c("Catharus fuscescens") #for single species plot

# x grid (your IQR-trimmed range)
vals <- unlist(get(covtoplot))
q <- quantile(vals, c(0.25, 0.75), na.rm = TRUE)
iqr <- q[2] - q[1]
vals <- vals[vals >= (q[1] - 1.5*iqr) & vals <= (q[2] + 1.5*iqr)]

newdata <- setNames(
  data.frame(seq(min(vals, na.rm = TRUE), max(vals, na.rm = TRUE), by = 1)),
  covtoplot
)

# predictions for the 9 species
pred_all <- bind_rows(lapply(sp_batch, function(sp) {
  p <- predict(covariate_results_df[[covtoplot]]$models[[sp]], newdata, "p")
  p[[covtoplot]] <- newdata[[covtoplot]]
  p$species <- sp
  p
}))

pred_all$species <- factor(pred_all$species, levels = unique(pred_all$species))

ggplot(pred_all, aes(x = .data[[covtoplot]], y = est)) +
  geom_line() +
  geom_line(aes(y = lower_0.95), linetype = "dashed") +
  geom_line(aes(y = upper_0.95), linetype = "dashed") +
  facet_wrap(~ species, ncol = 3) +
  labs(x = covtoplot, y = "Predicted detection") +
  lims(y = c(0, 1))

```
Plot distribution of a parameter across all birds for a covariate

```{r}
covtoplot  <- "radiationMin"
paramplot  <- "slope" 

vals <- sapply(combined_results, function(df) {
  df[covtoplot, paramplot]
})

ggplot(data.frame(value = vals), aes(x = value)) +
  geom_histogram(bins = 30, color = "black", fill = "grey80") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    x = paste(paramplot, "for", covtoplot),
    y = "Number of species"
  )
```

Let's plot factor models

```{r}
#events
data <- covariate_results_df$siteManagementEvents$models$`Buteo albonotatus`$real$p %>% 
  distinct(.keep_all = TRUE) %>%
  mutate(eventType = as.factor(c(
  "NA",
  "NoEvent",
  "populationSpike",
  "droughtPerceived",
  "grazing",
  "fire",
  "plantAddition",
  "otherNaturalDisturbance",
  "chemicalApplication",
  "plantReduction",
  "humanDisturbance"
)))

ggplot(data, aes(x = eventType, y = est)) +
  geom_bar(stat = 'identity') +
  geom_errorbar(aes(ymin = lower_0.95, ymax = upper_0.95), width = 0.2) +
  labs(x = "eventType",
       y = "Predicted Detection Probability") +
  lims(y = c(0, 1)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


# Various View statements
```{r}
View(combined_results$`Coragyps atratus`)

View(covariate_results_df$humidityMax$estimates$`Quiscalus major`)
View(covariate_results_df$humidityMax$detection_histories$`Chordeiles acutipennis`)

View(brd_joineddata_clean %>%
  filter(scientificName == "Catharus fuscescens"))
```

